A semantics of core AUTOSAR

### 2016-08-29 paper/paper.tex
Language polishing.

\documentclass[twocolumn]{article}
\usepackage{verbatim}
\usepackage[utf8x]{inputenc}
\usepackage[a4paper,margin=1.4cm]{geometry}
\usepackage[fleqn]{amsmath}
\usepackage{mathpartir}
\usepackage{amssymb}
\usepackage[makeroom]{cancel}
\usepackage{graphicx}

@@

@@

Components and ports do not carry any dynamic state besides the state of their constituent parts, so they need no explicit representation as process terms. However, in order to still be able to talk about model elements using their locally scoped names, we employ a hierarchical addressing scheme, such that address $i.r$ means runnable $r$ of component $i$ and $i.e$ means port element $e$ within component $i$. We use the following meta-variable conventions:

@@

The AUTOSAR specification restricts client-server connections to the many-to-one pattern only, assigning a dedicated runnable responsible for implementing the offered service. It furthermore forbids runnables to be triggered by any events other than operation invocations of compatible signatures, making it natural to associate the state needed to buffer up server invocations with the server runnables themselves. On the other hand, the specification also introduces timeouts and sequence counters that are private to each connection, which is why our calculus needs process terms that correspond to each required (i.e., client-side) operation as well. Axiom (33) shows how the current sequence counter $m$ blazes a call transition if the client-side operation is not already busy, while the busy case is detected in axiom (36). In axiom (34), a successful call transition causes the client identifier $a$ to be buffered up in the server runnable together with sequence number $m$ and call parameter $v$. Axiom (37) lets a server runnable ignore an aborted call attempt.

Here is an example of how a client runnable instance, a client port operation, and a server runnable interact during a call transition (assuming $\ToperationInvokedEvent b a$, $\Tconnect a {i.o}$, and $\TserverCallPointTimeout{i.o} t $).
$$
\begin{array}{c}
  {\Trinst {i.r} c \Xs {\TrteCall {p.o} v k}} \,\Opar  \\
  {\Toper {i.o} m {\Tdone \Tnodata}} \Opar \\
  {\Trunnable b 0 \Cs 0} \vspace{0.6ex} \\
  \red{\say{i.o}{\Tcall m v}} \\
  {\Trinst {i.r} c \Xs {\Tap k \Tok}} \,\Opar \\
  {\Toper {i.o} m {\Tcalling t}} \Opar \\
  {\Trunnable b 0 {\Tcons \Cs {\Tclient {i.o} m v}}  0}
\end{array}
$$

@@

A server runnable instance is identified by a client parameter distinct from $\Tvoid$. Axiom (42) says that when such a runnable instance reaches the end of its code sequence, it must emit the produced value together with the sequence number and \textbf{oper} address of its current invocation, before it is allowed to terminate (see Section~\ref{sect:SpawnTerm}). The addressed \textbf{oper} term reacts according to axiom (43), by completing the call, storing the produced value and increasing its sequence counter in preparation for the next call. However, should the \textbf{oper} timeout counter reach zero before a transition according to (43) can be taken, the client is obliged to unilaterally terminate the call and set the result to error code \textbf{timeout} (axiom (44)). Axiom (45) enables the completion of a call---successfully or via a timeout---to trigger a subscribing runnable (can be distinct from the original client).

Because of call timeouts, server runnables run the risk of producing results that the original client is not waiting for anymore. Detecting such mismatches is the job of sequence numbers, as witnessed by the requirement in axiom (43) that the sequence numbers returned by the server and expected by the client must be identical (variable $m$). However, there must also exist a means for server runnables to simply discard their results, otherwise they would not be able to terminate. Axiom (46) therefore defines an alternative transition for finished servers, applicable on the condition that the returned sequence number does \emph{not} match what the client expects (axiom (47)). By means of axiom (48), runnables can ignore the corresponding transition.

@@

So, in an abbreviated form, both the following derivations are possible:
$$
\begin{array}{l}
  q \Opar p \rightarrow \cdots \rightarrow q_1 \Opar p_1 \Opar p_{\text{inst}} \\
  q \Opar p \rightarrow \cdots \rightarrow q_1 \Opar p_2 \Opar p_{\text{inst}} \Opar p_{\text{inst}}
\end{array}
$$
That is, two data reception events in sequence can lead to the creation of either one or two runnable instances, and both behaviors are equally correct according to the AUTOSAR semantics.

@@

This anomaly may lead to the conclusion that the AUTOSAR specification does not really intend more than one timing event to be defined per runnable. Luckily, our formalization is also compatible with this interpretation, as the restriction only concerns what set of static models to expect as starting states, not their dynamic behavior.

Closely related to the interleaving of timing events, and of interest even if only distinct runnables are being triggered, is the notion of initial timer offsets. That is, at what time will each timer tick for the first time --- immediately at startup or at some other (later) time? The AUTOSAR specification only introduces the notion of timing offset in relation to the operating system tasks that are expected to implement runnables; the Software Component Template itself is totally silent on the issue \cite{AR:SWC}. Lacking an abstract definition of runnable offsets, we have chosen to simply fire all timers simultaneously at $0$ seconds from startup (hence the value $0$ in the initial {\bf timer} term definition in Section~\ref{sec:Sem}). It is not hard to see that other values may be equally reasonable, though, like
$$
	\Ttimer {i.r} {t_p} {t_p}
$$
which lets each timer wait one full cycle before triggering for the first time. However, should we really want to take the assumed implementation offsets into account, the initial timer state would rather be
$$
	\Ttimer {i.r} {t_p} {t_o}
$$
for some chosen non-negative $t_o$. One might even want to add the condition $t_o < t_p$, but it must be emphasized that all these initilization options are speculative --- the AUTOSAR Software Component Template must be clarified before a sufficiently abstract model of timed behavior can be obtained.

@@

Another concern regarding exclusive areas is that although the specification clearly demands that exiting is done in the reverse order of entering, it does not provide any means for the exit command to report detected violations of that condition \cite[ch.~5.6.29]{AR:RTE}. This means that if a runnable instance tries to exit exclusive area $x$, and $x$ is not the latest exclusive area acquired by that instance, the operation is neither allowed to succeed nor able to flag an error due to a \emph{void} return type. This leaves our transition semantics with no other option than to indefinitely block progress for such a runnable, which is the consequence of providing no transition alternative besides axiom (11) to a runnable instance about to execute the exit command. Had the specification allowed for an error result, however, the following axioms could have been added to resolve the order violation cases:
$$
\begin{array}{lcll}
		{\Trinst {i.r} c {\Tcons y \Xs} {\TrteExit x k}}
		\;\; \red{\Tsay {i.r} \Tnop} \\
		{\Trinst {i.r} c {\Tcons y \Xs} {\Tap k {\kw{orderErr}}}}
		& \Pif\;x \neq y
\\ \\
		{\Trinst {i.r} c \Teps {\TrteExit x k}}
		\;\; \red{\Tsay {i.r} \Tnop} \\
		{\Trinst {i.r} c \Teps {\Tap k {\kw{orderErr}}}}

@@

A transition semantics like the one defined in this report provides rules for deriving the correct behaviors, while leaving the incorrect behaviors without derivation alternatives. This has the effect that a process term which attempts to do something illegal is assigned no meaningful remaining trace; it appears indefinitely stuck in its current state as if it were deadlocked and can participate in subsequent transitions only by ignoring what is being said.

@@

First the code syntax needs to be extended so that failing computations can be represented:
$$
\begin{array}{lll}
  \Code & ::= & \ldots               \\
        & |   & \Return v            \\
        & |   & \Throw
\end{array}
$$
A "throw" command has no continuation and is thus similar to the exception generating commands of many modern programming languages. It could be given an informative value parameter as well, but this will not be needed in this simplistic proposal. Throw commands can appear explicitly in the code to indicate software-detected errors, or they can be the result of special failure transitions, like the following alternative resolution of an exit order violation:
$$
\begin{array}{lcll}
		{\Trinst {i.r} c {\Tcons y \Xs} {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail} \\
		{\Trinst {i.r} c {\Tcons y \Xs} \Throw}
		& \Pif\;x \neq y
\\ \\
		{\Trinst {i.r} c \Teps {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail}  \\
		{\Trinst {i.r} c \Teps \Throw}
\end{array}
$$
Because a failed runnable instance now has an explicit representation, we can specify details of how such a process should interact with its enviroment. We may for example wish to define that a special error code is returned if our runnable is running on behalf of a client, and that any exclusive areas held are automatically exited in the right order (c.f.~axioms (42) and (11)):
$$
\begin{array}{lcll}
		{\Trinst a {\Tclient b m \_} \Xs \Throw}
		\;\; \red{\Tsay b {\Tret m {\kw{serverErr}}}} \\
		{\Trinst a \Tvoid \Xs \Throw}
\\ \\
		{\Trinst {i.r} \Tvoid {\Tcons x \Xs} \Throw}
		\;\; \red{\Tsay {i.x} \Tex} \\
		{\Trinst {i.r} \Tvoid \Xs \Throw}
\end{array}
$$
Finally, we can ensure that the instance counter of a runnable is properly decremented even if an instance terminates due to a failure (in interaction with axiom (52)):
$$
\begin{array}{rcll}
		{\Trinst a \Tvoid \Teps \Throw}
		& \red{\Tsay a \Tterm}
		\Tzero
\end{array}
$$
Other design choices are of course perfectly possible, and more elaborate rules, for example involving an explicit exception-catching mechanism, could be considered as well. We must emphasize, though, that the rules in this section are just hypothetical examples; they are not based on the current AUTOSAR specification and are only presented to illustrate the succinctness and level of detail that can be obtained by a formal semantics aproach.

>>>>

\documentclass[10pt,conference]{IEEEtran}
%\documentclass[twocolumn]{article}
\usepackage{verbatim}
\usepackage[utf8x]{inputenc}
\usepackage[a4paper,margin=1.4cm]{geometry}
\usepackage[fleqn]{amsmath}
\usepackage{mathpartir}
\usepackage{amssymb}
\usepackage[makeroom]{cancel}
\usepackage{graphicx}

@@

%TODO: Some simple (informal) example early to give the reader an early view of flavour of the results.
%  Suggestion: use something like the "two values sent to a queue with a trigger runnable can lead to the runnable triggering once or twice" example

@@

Components do not carry any dynamic state besides the state of their constituent parts, so they need no explicit representation as process terms. However, in order to still be able to talk about model elements using their locally scoped names, we employ a hierarchical addressing scheme, such that address $i.r$ means runnable $r$ of component $i$ and $i.e$ means port element $e$ within component $i$. We use the following meta-variable conventions:

@@

The AUTOSAR specification restricts client-server connections to the many-to-one pattern only, assigning a dedicated runnable responsible for implementing the offered service. It furthermore forbids runnables to be triggered by any events other than operation invocations of compatible signatures, making it natural to associate the state needed to buffer up server invocations with the server runnables themselves. On the other hand, the specification also introduces timeouts and sequence counters that are private to each connection, which is why our calculus also needs process terms that correspond to each required (i.e., client-side) operation. Axiom (33) shows how the current sequence counter $m$ blazes a call transition if the client-side operation is not already busy, while the busy case is detected in axiom (36). In axiom (34), a successful call transition causes the client identifier $a$ to be buffered up in the server runnable together with sequence number $m$ and call parameter $v$. Axiom (37) lets a server runnable ignore an aborted call attempt.

Here is an example of how a client runnable instance, a client port operation and a server runnable interact during a call transition (assuming $\ToperationInvokedEvent b a$, $\Tconnect a {i.o}$, and $\TserverCallPointTimeout{i.o} t $).
$$
\begin{array}{c}
  {\Trinst {i.r} c \Xs {\TrteCall {p.o} v k}} \,\Opar  \\
  {\Toper {i.o} m {\Tdone \Tnodata}} \Opar \\
  {\Trunnable b 0 \Cs 0} \vspace{0.6ex} \\
  \red{\say{i.o}{\Tcall m v}} \\
  {\Trinst {i.r} c \Xs {\Tap k \Tok}} \,\Opar \\
  {\Toper {i.o} m {\Tcalling t}} \Opar \\
  {\Trunnable b 0 {\Tcons \Cs {\Tclient {i.o} m v}}  0}
\end{array}
$$

@@

A server runnable instance is identified by a client parameter distinct from $\Tvoid$. Axiom (42) says that when such a runnable instance reaches the end of its code sequence, it must announce the produced value, together with the sequence number and \textbf{oper} address of its current invocation, before it is allowed to terminate (to be defined in Section~\ref{sect:SpawnTerm}). The addressed \textbf{oper} term reacts according to axiom (43), by completing the call, storing the produced value and increasing its sequence counter in preparation for the next call. However, should the \textbf{oper} timeout counter reach zero before a transition according to (43) can be taken, the client is obliged to unilaterally terminate the call and set the result to error code \textbf{timeout} (axiom (44)). Axiom (45) enables the completion of a call---successfully or via a timeout---to trigger a subscribing runnable (can be distinct from the original client).

Because of call timeouts, server runnables run the risk of producing results that the original client is not waiting for anymore. Detecting such mismatches is the job of the sequence numbers, as witnessed by the requirement in axiom (43): the sequence number returned by the server and the one expected by the client must be identical (variable $m$). However, there must also exist a means for server runnables to simply discard their results, otherwise they would not be able to terminate. Axiom (46) therefore defines an alternative transition for finished servers, applicable on the condition that the returned sequence number does \emph{not} match what the client expects (axiom (47)). By means of axiom (48), subscribing runnables are allowed to ignore the corresponding transition.

@@

So, in an abbreviated form, both the following derivations are possible for the same reception event sequence:
$$
\begin{array}{l}
  q \Opar p \rightarrow \cdots \rightarrow q_1 \Opar p_1 \Opar p_{\text{inst}} \\
  q \Opar p \rightarrow \cdots \rightarrow q_1 \Opar p_2 \Opar p_{\text{inst}} \Opar p_{\text{inst}}
\end{array}
$$
That is, the system can respond by creating either one or two runnable instances, and both behaviors are equally correct according to the AUTOSAR semantics.

@@

This anomaly may lead to the conclusion that the AUTOSAR specification does not really intend to allow more than one timing event per runnable. Luckily, our formalization is also compatible with this interpretation, as the restriction only concerns what set of static models to expect as starting states, not their dynamic behavior.

Closely related to the interleaving of timing events, and of interest even if only distinct runnables are being triggered, is the notion of initial timer offsets. That is, at what time will each timer tick for the first time --- immediately at startup or at some other (later) time? The AUTOSAR specification only introduces a concept of timing offsets in relation to the operating system tasks that are expected to implement runnables; the Software Component Template itself is totally silent on the issue \cite{AR:SWC}. Lacking an abstract definition of runnable offsets, we have chosen to simply fire all timers simultaneously at $0$ seconds from startup (hence the value $0$ in the initial {\bf timer} term definition in Section~\ref{sec:Sem}). It is not hard to see that other values may be equally reasonable, though, like
$$
	\Ttimer {i.r} {t_p} {t_p}
$$
which lets each timer wait one full cycle before triggering for the first time. However, should we really want to take the assumed task offsets into account, the initial timer state would rather be
$$
	\Ttimer {i.r} {t_p} {t_o}
$$
for some non-negative $t_o$. One might even want to add the condition $t_o < t_p$, but it must be emphasized that all these initilization options are speculative --- the AUTOSAR Software Component Template must be clarified before a sufficiently abstract model of timed behavior can be obtained.

@@

Another concern regarding exclusive areas is that although the specification clearly demands that exiting is done in the reverse order of entering, it does not provide any means for the exit command to report detected violations of that condition \cite[ch.~5.6.29]{AR:RTE}. This means that if a runnable instance tries to exit exclusive area $x$, and $x$ is not the latest exclusive area acquired by that instance, the operation is neither allowed to succeed, nor able to flag an error (due to a \emph{void} return type). This leaves our transition semantics with no other option than to indefinitely block progress for such a runnable, which is the consequence of providing no transition alternative besides axiom (11) to runnable instances about to execute the exit command. Had the specification allowed for an error result, however, the following axioms could have been added to resolve the order violation cases:
$$
\begin{array}{lcll}
		{\Trinst {i.r} c {\Tcons y \Xs} {\TrteExit x k}}
		\;\; \red{\Tsay {i.r} \Tnop} \\
		{\Trinst {i.r} c {\Tcons y \Xs} {\Tap k {\kw{orderErr}}}}
		& \Pif\;x \neq y
\\ \\
		{\Trinst {i.r} c \Teps {\TrteExit x k}}
		\;\; \red{\Tsay {i.r} \Tnop} \\
		{\Trinst {i.r} c \Teps {\Tap k {\kw{orderErr}}}}

@@

A transition semantics like the one defined in this report provides rules for deriving the correct behaviors, while leaving the incorrect behaviors without derivation alternatives. This has the effect that a process term which attempts to do something illegal is assigned no meaningful remaining trace; it appears indefinitely stuck in its current state as if it were deadlocked, and can participate in subsequent transitions only by ignoring what is being said.

@@

First the code syntax needs to be extended so that failing computations can be represented:
$$
\begin{array}{lll}
  \Code & ::= & \ldots               \\
        & |   & \Return v            \\
        & |   & \Throw
\end{array}
$$
A "throw" command has no continuation and is thus similar to the exception generating commands of many modern programming languages. It could be given an informative value parameter as well, but this will not be needed in our simplistic proposal. Throw commands can appear explicitly in the code to indicate software-detected errors, or they can be the result of special failure transitions, like the following alternative resolution of an exit order violation:
$$
\begin{array}{lcll}
		{\Trinst {i.r} c {\Tcons y \Xs} {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail} \\
		{\Trinst {i.r} c {\Tcons y \Xs} \Throw}
		& \Pif\;x \neq y
\\ \\
		{\Trinst {i.r} c \Teps {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail}  \\
		{\Trinst {i.r} c \Teps \Throw}
\end{array}
$$
Because a failed runnable instance now has an explicit representation, we can specify details of how such a process should interact with its enviroment. We may for example wish to define that a special error code is returned if our runnable is running on behalf of a client, and that any exclusive areas held are automatically exited in the right order (c.f.~axioms (42) and (11)):
$$
\begin{array}{lcll}
		{\Trinst a {\Tclient b m \_} \Xs \Throw}
		\;\; \red{\Tsay b {\Tret m {\kw{serverErr}}}} \\
		{\Trinst a \Tvoid \Xs \Throw}
\\ \\
		{\Trinst {i.r} \Tvoid {\Tcons x \Xs} \Throw}
		\;\; \red{\Tsay {i.x} \Tex} \\
		{\Trinst {i.r} \Tvoid \Xs \Throw}
\end{array}
$$
Finally, we can ensure that the instance counter of a runnable is properly decremented even if an instance terminates due to a failure (in interaction with axiom (52)):
$$
\begin{array}{rcll}
		{\Trinst a \Tvoid \Teps \Throw}
		& \red{\Tsay a \Tterm}
		\Tzero
\end{array}
$$
Other design choices are of course perfectly possible, and more elaborate rules, for example involving an explicit exception-catching mechanism, could be considered as well. We must emphasize, though, that the rules in this section are just hypothetical examples; they are not based on the current AUTOSAR specification and are only presented to illustrate the succinctness and level of detail that can be obtained by a formal semantics aproach.
